---
title: "DSEMrand: Full Workflow and Sensitivity Analysis"
author: "Joe Caracappa"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DSEMrand: Full Workflow and Sensitivity Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 6
)
library(igraph)
library(DSEMrand)
```

## Introduction

This vignette demonstrates the complete end-to-end workflow of the DSEMrand package. We will walk through simulating state-space ecosystem data from a hypothesized Directed Acyclic Graph (DAG), fitting a dynamic structural equation model (DSEM), and rigorously validating the structural topology using Monte Carlo randomizations and automated sensitivity testing.

### The Reference Ecosystem Model

First, we load our hypothesized ecosystem topology and visualize it. This reference DAG defines the assumed causal pathways and temporal lags in our system.

```{r}
# 1. Generate the reference matrix
ref_adj <- read.csv(here::here('data-raw', 'example_weight_dag.csv'), row.names = 1) |> as.matrix()
ref_adj[which(ref_adj != 0)] <- 1

# 2. Convert it to an igraph object
p <- igraph::graph_from_adjacency_matrix(ref_adj, mode = "directed")

# 3. Plot using a hierarchical layout
igraph::plot.igraph(p, 
                    layout = igraph::layout_with_sugiyama(p)$layout, 
                    vertex.color = "lightblue", 
                    vertex.size = 30, 
                    edge.arrow.size = 0.6,
                    main = "Reference Hierarchical DAG")
```

### 2. Simulating Time Series Data

To test our methodology, we simulate a robust dataset containing 300 time-steps. We apply varying levels of baseline drift, noise, and define a latent "food quality" construct measured by three physical/biological indicators.

```{r}
dsem.data <- DSEMrand::simulate_dsem_data(
  weights.file = here::here('data-raw', 'example_weight_dag.csv'),
  lags.file = here::here('data-raw', 'example_lags_dag.csv'),
  latent_vars = 'food.quality',
  n_steps = 300,
  var.mu = c('bot.temp' = 5, 'diatom' = 1E6, 'juv.bio' = 1E5, 'adult.bio' = 1E4, 'catch' = 1E3, 'food.quality' = 100, 'growth' = 10, 'stratification' = 10, 'diatom.prop' = 0.7),
  var.sd = c('bot.temp' = 1, 'diatom' = 1E3, 'juv.bio' = 1E3, 'adult.bio' = 1E3, 'catch' = 1E2, 'food.quality' = 10, 'growth' = 3, 'stratification' = 3, 'diatom.prop' = 0.05),
  var.slope = c('bot.temp' = 0.025, 'diatom' = 0, 'juv.bio' = -0.01, 'adult.bio' = -0.05, 'catch' = 0, 'food.quality' = 0.025, 'growth' = 0.05, 'stratification' = 0.01, 'diatom.prop' = 0),
  diagnostics = TRUE
)

# Print diagnostic summary showing true correlation footprints
print(dsem.data$diagnostic_summary)

# Save for downstream modeling
saveRDS(dsem.data$data, here::here('data-raw', 'dsem_example_data.rds'))
```

### 3. Fitting the Generalized DSEM

Next, we run the simulated, raw-scale data through our automated DSEM solver. The function automatically detrends, standardizes, and anchors our 3-indicator latent complex to ensure strict mathematical identifiability.

```{r}
dsem.data.test <- DSEMrand::fit_generalized_dsem(
  adj.file = here::here('data-raw', 'example_unknown_dag.csv'),
  lags.file = here::here('data-raw', 'example_lags_dag.csv'),
  data.file = here::here('data-raw', 'dsem_example_data.rds'),
  latent_dict = list('food.quality' = c('growth', 'stratification', 'diatom.prop')),
  detrend = TRUE, 
  standardize = TRUE, 
  control_time_drift = FALSE
)

# Display the recovered structural estimates
print(dsem.data.test$estimates)

saveRDS(dsem.data.test, here::here('data-raw', 'dsem_example_results.rds'))
```

We can compare these estimates back to our known ground-truth to verify that the estimator successfully unpacked the complex topology.

```{r}
comparison_table <- DSEMrand::compare_dsem_results(
  true_weights_file = here::here('data-raw', 'example_weight_dag.csv'),
  dsem_output = readRDS(here::here('data-raw', 'dsem_example_results.rds'))
)
print(comparison_table)
```

### 4. Network Randomization

To prove our results aren't just spurious correlations, we generate empirical null models by randomizing the network edges. Importantly, the randomization safely preserves the measurement structure of the food.quality latent complex.

```{r}
ref_adj_sign <- read.csv(here::here('data-raw', 'example_weight_dag.csv'), row.names = 1) |> as.matrix() |> sign()

rand_adj <- DSEMrand::randomize_dsem_graph(
  ref_adj_sign,
  ref_lags = here::here('data-raw', 'example_lags_dag.csv'),
  latent_dict = list('food.quality' = c('growth', 'stratification', 'diatom.prop'))
)

# Plot one of the randomized topologies
g <- igraph::graph_from_adjacency_matrix(rand_adj$adj, mode = "directed")
igraph::plot.igraph(g, 
                    layout = igraph::layout_with_sugiyama(g)$layout, 
                    vertex.color = "lightblue", 
                    vertex.size = 30, 
                    edge.arrow.size = 0.6,
                    main = "Randomized Hierarchical DAG")
```

### 5. Monte Carlo Structural Robustness

We generate a batch of 100 unique random topologies and fit the DSEM model to every single one to create an empirical null distribution for our Negative Log-Likelihood (NLL) and our parameter estimates.

```{r}
# 1. Generate 100 unique valid graphs
batch_results <- DSEMrand::evaluate_random_graphs(
  ref_adj_sign,
  ref_lags = here::here('data-raw', 'example_lags_dag.csv'),
  N = 100, 
  deduplicate = TRUE,
  latent_dict = list('food.quality' = c('growth', 'stratification', 'diatom.prop'))
)

# 2. Fit the data to all 100 graphs in parallel
batch_fits <- DSEMrand::fit_random_graphs(
  batch_results = batch_results,
  data.file = here::here('data-raw', 'dsem_example_data.rds'),
  cores = 8,
  latent_dict = list('food.quality' = c('growth', 'stratification', 'diatom.prop'))
)

# 3. Compare the Reference fit against the Random Batch
results <- DSEMrand::compare_dsem_distributions(
  ref_fit = dsem.data.test, 
  batch_fits, 
  plot = TRUE
)

saveRDS(results,here::here('data','dsem_randomization_results.rds'))
```

```{R}

results = readRDS(here::here('data','dsem_randomization_results.rds'))
# Visualize the overall structural dependency
print(results$plot)

# Review which specific pathways are highly dependent on the reference topology
print(results$model_fit_comparison)
print(results$parameter_comparison)
```

### 6. Sensitivity Analysis

Finally, we stress-test the methodology by systematically degrading the data. We sequentially introduce missing data points, spike the background noise (variance), and shrink the time-series length to identify the exact thresholds where our reference topology fails to outperform random noise.

```{r, message = T, warning = F}
# Run the automated sensitivity workflow
# sens_results <- DSEMrand::run_dsem_sensitivity(
#   adj_file = here::here('data-raw', 'example_unknown_dag.csv'),
#   lags_file = here::here('data-raw', 'example_lags_dag.csv'),
#   node_classes = list(
#     source = c("bot.temp", "diatom"),
#     intermediate = c("juv.bio", "adult.bio"),
#     sink = c("catch"),
#     measurement = c("growth", "stratification", "diatom.prop")
#   ),
#   latent_dict = list('food.quality' = c('growth', 'stratification', 'diatom.prop')),
#   base_mu = c('bot.temp'=5, 'diatom'=1E6, 'juv.bio'=1E5, 'adult.bio'=1E4, 'catch'=1E3, 'food.quality'=100, 'growth'=10, 'stratification'=10, 'diatom.prop'=0.7),
#   base_sd = c('bot.temp'=1, 'diatom'=1E3, 'juv.bio'=1E3, 'adult.bio'=1E3, 'catch'=1E2, 'food.quality'=10, 'growth'=3, 'stratification'=3, 'diatom.prop'=0.05),
#   base_slope = c('bot.temp'=0.025, 'diatom'=0, 'juv.bio'=-0.01, 'adult.bio'=-0.05, 'catch'=0, 'food.quality'=0.025, 'growth'=0.05, 'stratification'=0.01, 'diatom.prop'=0),
#   out_dir = here::here('data'),
#   cores = 6,
#   N_rand = 5,
#   M_preflight = 5,
#   max_ts_length = 400
# )
# saveRDS(sens_results, here::here('data','sensitivity_results.rds'))
```

```{r}
sens_results = readRDS(here::here('data','sensitivity_results_all.rds'))
# 1. Inspect Baseline Robustness (Pre-Flight Check)

print(sens_results$preflight_summary)
print(sens_results$plot_preflight)

# 2. Inspect Degradation Trajectories
print(sens_results$plot_sensitivity)
```

